---
title: "Data handling for anticlustering"
output: rmarkdown::html_vignette
author: Martin Papenberg
vignette: >
  %\VignetteIndexEntry{Anticlustering_in_2025}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)

```

```{r setup}
library(anticlust)
```

This vignette is (in September 2025) the most up-to-date reference for doing anticlustering. Here, I summarize some of the things I learned about---or implemented for---anticlustering, which may not be documented in earlier work (e.g., papers, other vignettes). 

To reproduce the following example, install the "MASS" package that contains the `survey` data set that is used.

```{r}
library(MASS)
data(survey)       # load data set
nrow(survey)       # number of students
head(survey, n=10) # look at the data
``` 

The `survey` data set contains some demographic information on a student sample. Imagine you wish to assign students to three different dormitories and want to create a similar groups of students in each house. In this example I will use the following features (brackets contain the description contained in the R documentation, `?MASS::survey`):

**Categorical features:**

- `Sex`: The sex of the student (Factor with levels "Male" and "Female".)
- `W.Hnd`: writing hand of student. (Factor, with levels "Left" and "Right".)
- `Exer`: how often the student exercises. (Factor, with levels "Freq" (frequently), "Some", "None".)
- `Smoke`: how much the student smokes. (Factor, levels "Heavy", "Regul" (regularly), "Occas" (occasionally), "Never".)


**Categorical features:**

- `Pulse`: pulse rate of student (beats per minute).
- `Height`: height of the student in centimetres.
- `Age`: age of the student in years.

I will select these features as two separate data frames, one for categorical features and one for the numeric features. 

```{r}

features <- c("Sex", "W.Hnd", "Exer", "Smoke", "Pulse", "Height", "Age")

survey$House <- anticlustering(survey[, features], K = 3, objective = "kplus", standardize = TRUE, method = "3phase")

library(tableone)

CreateTableOne(features, strata = "House", data = survey)

```
