---
title: "Using the R package anticlust for stimulus selection"
output: 
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# define a method for objects of the class data.frame
knit_print.matrix = function(x, ...) {
    res = paste(c("", "", kable(x, row.names = TRUE)), collapse = "\n")
    asis_output(res)
}
# register the method
registerS3method("knit_print", "matrix", knit_print.matrix)

```

This document was last updated on `r Sys.Date()`. **It is still under 
active construction.**

This vignette uses functionality that is currently only available in the 
development version of the `anticlust` package. If you would 
like to check out the latest features to follow this tutorial, install 
the development version by executing the following code:

```{r, eval = FALSE}
if (!"remotes" %in% rownames(installed.packages())) {
  install.packages("remotes")
}
library(remotes)
install_github("m-Py/anticlust@devel")
```

Now, load the package to reproduce the code in this tutorial:

```{r setup}
library(anticlust)
```

---

This tutorial teaches you how to use the `R` package `anticlust` for 
stimulus selection. We use norming data for the open stimulus 
pool OASIS for this purpose (Brielmann & Pelli, 2019; Kurdi, Lozano, & 
Banaji, 2017). We load the data provided by Brielmann and Pelli (2019)
from the internet and save the data as an `R` data frame: 

```{r}
oasis <- read.csv("https://raw.githubusercontent.com/aenneb/OASIS-beauty/master/means_per_image.csv")
```

Let's check out the data: 

```{r}
head(oasis)
nrow(oasis)
```

The OASIS data set contains norming data for 900 photos on 
arousal, valence, and beauty; so we are interested in the columns 
`Valence_mean`, `Arousal_mean`, `beauty_mean` to select stimulus sets.

In this tutorial, we get to know at four different procedures. The four 
procedures result by crossing two properties that characterize the
selection of stimuli:

- The final pool of stimuli that is part of the experiment is... 
    + ... manually selected by the researcher; this selection is then
      automatically partitioned into subsets using anticlustering
    + ... selected by a computer -- the computer does all of the work
- The experimental manipulation is ...
    + ... realized via the intrinsic properties of the stimuli 
    + ... independent of the properties of the stimuli

# The researcher decides

First, I discuss experimental designs where the final set of stimuli has 
already been selected by the researcher. Now, this selection has to be 
divided into several experimental sets. I distinguish two cases: 

1. The stimulus sets have to be as similar as possible with regard to 
*all* response-relevant variables.

2. The groups have to differ on an independent variable, but should be 
as similar as possible on other variables. 

## Experimental manipulation independent of stimuli: Anticlustering

When employing an experimental manipulation that is orthogonal to the 
stimuli, we can use anticlustering to partition the stimulus pool into 
equal-sized parts that are as similar as possible. In particular, we 
employ *anticluster editing* to assign stimuli to sets (see Papenberg 
and Klau, 2019).

For this example, I preselect 200 images, namely all images that are 
identified as "objects" in the data base.

```{r}
objects <- oasis[oasis$Category == "Object", ]
```

Now I create four sets from the pool of 200 stimuli. The four sets sets 
should be as similar as possible with regard to the three variables 
arousal, valence, and beauty. I use the function `select_stimuli()` that 
internally calls the `anticlustering()` function (also see 
`?anticlustering`).

```{r}
# Create character vector with the relevant covariates:
covariates <- c("Valence_mean", "Arousal_mean", "beauty_mean")
stimuli <- select_stimuli(
  objects,
  equalize = covariates,
  design = 4 # specifies the number of groups
)
```

The output variable `stimuli` is a data frame that has the same columns 
as the original input (the data frame `objects`), but has an additional 
column called `SET` that encodes the stimulus set that each item was 
assigned to.

We can use the `anticlust` function `mean_sd_tab()` to check out the 
quality of the results. This function prints the means and standard 
deviations (in brackets) of the variables of interest for each of the 
four stimulus sets:

```{r}
mean_sd_tab(stimuli[, covariates], stimuli$SET)
```

Apparently, the stimulus assignment worked quite well.

## Experimental manipulation realized via stimuli: Min-Max Anticlustering

In a second application, we create three sets that are dissimilar with 
regard to valence (i.e., valence would be our independent variable in an 
experiment), but similar with regard to arousal. In this example, we use 
the 134 images that are classified as "animals": 

```{r}

animals <- oasis[oasis$Category == "Animal", ]
stimuli <- select_stimuli(
  animals,
  split_by = "Valence_mean",
  equalize = c("Arousal_mean"),
  design = 2 # specifies the number of groups
)

mean_sd_tab(stimuli[, c("Valence_mean", "Arousal_mean")], stimuli$SET)

```

This worked quite well. Note that the task is significantly more 
difficult when we also include beauty as a covariate:

```{r}

stimuli <- select_stimuli(
  animals,
  split_by = "Valence_mean",
  equalize = c("Arousal_mean", "beauty_mean"),
  design = 2 # specifies the number of groups
)

mean_sd_tab(
  stimuli[, c("Valence_mean", "Arousal_mean", "beauty_mean")], 
  stimuli$SET
)

```

The sets are no longer very dissimilar with regard to valence. The 
problem is that valence is strongly correlated with beauty, as shown in 
the following scatter plot: 

```{r, fig.width = 4.5, fig.height = 4.5}
plot(
  animals[, c("Valence_mean", "beauty_mean")], 
  pch = 4, 
  cex = 0.7, 
  col = c("#a9a9a9", "orange")[stimuli$SET]
)
cor(animals$Valence_mean, animals$beauty_mean)
```

It is more or less impossible to create sets that differ strongly on 
valence but are similar with regard to beauty, because these variables 
covary too much. Selecting item sets from a larger item pool -- rather 
than just partitioning a preselected pool into parts -- helps to ensure 
that the sets are more dissimilar with regard to the independent 
variable. This application is discussed in the next section.

# The computer does all the work

In this section we discuss the case when item sets are extracted 
from the entire stimulus pool in an entirely automated manner -- no 
human input required.

## Experimental manipulation independent of stimuli: Subset anticlustering

Imagine I want to select 3 sets of stimuli that are similar with regard 
to all numeric variables. 

```{r}
stimuli <- select_stimuli(
  oasis,
  equalize = c("Valence_mean", "Arousal_mean", "beauty_mean"),
  design = 3, # specifies the number of groups
  n = 100 # Number of stimuli per set
)
```

To indicate that I do not want to partition the entire stimulus pool, 
but only extract a subset of the pool, I used the argument `n` -- which 
is the size of **each** stimulus set. As the `nrow()` command confirms, 
our selection consists of 300 stimuli:

```{r}
nrow(stimuli)
```

Inspecting means and standard deviations of the variables by set shows 
that the stimulus assignment was successful:

```{r}
mean_sd_tab(
  stimuli[, c("Valence_mean", "Arousal_mean", "beauty_mean")], 
  stimuli$SET
)
```

Hard to beat those results.

## Experimental manipulation realized via stimuli: Subset Min-Max-Anticlustering

In this section, I select item sets that differ with regard to valence,
but are similar with regard to arousal and beauty. I select 3 sets รก 40 
items from the entire pool of 900 items:

```{r}

stimuli <- select_stimuli(
  oasis,
  split_by = "Valence_mean",
  equalize = c("Arousal_mean", "beauty_mean"),
  design = 3, # specifies the number of groups
  n = 40 # Number of stimuli per set
)

mean_sd_tab(
  stimuli[, c("Valence_mean", "Arousal_mean", "beauty_mean")], 
  stimuli$SET
)
```

Visualizing this selection is interesting. The following panel 
illustrates the pairwise scatter plots:

```{r, fig.width = 7, fig.height = 7}
# Illustrate final stimulus selection
was_selected <- oasis$item %in% stimuli$item
group <- rep(1, nrow(oasis))
group[was_selected] <- stimuli$SET + 1


colors <- c("#a9a9a9", "#3cb44b", "#f032e6", "#093145")
cex <-  c(0.1, rep(1.1, 3))
pch <- c(4, 15, 17, 19)
plot(
  oasis[, c("Valence_mean", "Arousal_mean", "beauty_mean")],
  col = colors[group],
  cex = cex[group],
  pch = pch[group]
)
```

Small grey symbols indicate that an image has not been selected; large 
colored dots illustrate images that have been selected. We recognize the 
following properties of our selection:

- Even thought it is very difficult to obtain sets that differ on 
valence but are similar on beauty -- the two variables are strongly 
correlated -- the program did it (see the scatter plot of beauty and
valence in the lower left or upper right)
- There is actually overlap in valence between the sets. The neutral 
condition also contains some very positive and very negative images. 
Using a Divide and Select approach (see below), there would 
not have been any overlap on valence
- The three sets overlap strongly on the variables beauty and arousal,
as they should

## Divide and Select

The `divide_and_select()` function can also be used to create groups 
that differ on an independent variable. It works by actually dividing
the item pool on this variable using a hard cutoff. It works as follows:

```{r}

stimuli <- divide_and_select(
  oasis,
  split_by = "Valence_mean",
  equalize = c("Arousal_mean", "beauty_mean"),
  design = 3, # specifies the number of groups
  n = 40 # Number of stimuli per set
)

mean_sd_tab(
  stimuli[, c("Valence_mean", "Arousal_mean", "beauty_mean")], 
  stimuli$SET
)

```

Let's also visualize this selection:

```{r, fig.width = 7, fig.height = 7}
was_selected <- oasis$item %in% stimuli$item
group <- rep(1, nrow(oasis))
group[was_selected] <- stimuli$SET + 1

plot(
  oasis[, c("Valence_mean", "Arousal_mean", "beauty_mean")],
  col = colors[group],
  cex = cex[group],
  pch = pch[group]
)
```

With the Divide and Select method, there is no overlap in valence 
between the different sets -- they were actually "divided" by valence. 
However, in total, the dissimilarity with regard to valence (based on 
the mean difference between) is lower than in the example above. Hence, 
I would suggest to try out the Min-Max-Anticlustering approach first 
(using the function `select_stimuli()`) as it tends to better maximize 
differences between the `split` variable while also ensuring that the 
covariates are balanced. However, the choice is entirely left to you;
it is probably best to try out both methods. 

### Divide by categorical variable

In the previous example, valence was artificially categorized to serve
as a grouping variable. Sometimes, a truly categorical variable should
serve a the grouping variable. As an example, each of the images in the 
data base belongs to one of four categories. The frequency of each 
category is given as follows:

```{r}
table(oasis$Category)
```

Imagine I want to select 30 images from each category and the four
sets should be as similar as possible with regard to arousal and
valence. We can use the function `select_stimuli()` for this purpose:

```{r}

stimuli <- divide_and_select(
  oasis,
  split_by = "Category",
  equalize = c("Arousal_mean", "Valence_mean"),
  design = 4, # specifies the number of groups
  n = 40 # Number of stimuli per set
)

mean_sd_tab(stimuli[, c("Arousal_mean", "Valence_mean")], stimuli$SET)
```

Apparently, this worked quite well. Let's also confirm that the 
`Category` variable is cleanly split between the four sets:

```{r}
table(stimuli$Category, stimuli$SET)
```

### Divide and Select Reloaded

Using the functionality to pass categorical variables to the 
`divide_and_select()` function, we can try to select item sets that
differ more strongly with regard to an independent variable. To this
end, I apply customized cutoffs above (and below) which the valence
levels are defined. I use the function `case_when()` from the package
`dplyr`:

```{r}

library(dplyr)

oasis$category_valence <- case_when(
  oasis$Valence_mean < 3 ~ "low",
  oasis$Valence_mean > 3.5 & oasis$Valence_mean < 4.5 ~ "neutral",
  oasis$Valence_mean > 5 ~ "high",
  TRUE ~ "" # default case: exclude those
)

# exclude cases that are not within one of the ranges
oasis_subset <- oasis[oasis$category_valence != "", ]

# Check out number of preselected items:
nrow(oasis_subset)

# Do Divide and Select
stimuli <- divide_and_select(
  oasis_subset,
  split_by = "category_valence",
  equalize = c("Arousal_mean", "beauty_mean"),
  design = 3, # specifies the number of groups
  n = 40 # Number of stimuli per set
)

mean_sd_tab(
  stimuli[, c("Valence_mean", "Arousal_mean", "beauty_mean")], 
  stimuli$SET
)
```

Let's also visualize this selection:

```{r, fig.width = 7, fig.height = 7}
was_selected <- oasis$item %in% stimuli$item
group <- rep(1, nrow(oasis))
group[was_selected] <- stimuli$SET + 1

plot(
  oasis[, c("Valence_mean", "Arousal_mean", "beauty_mean")],
  col = colors[group],
  cex = cex[group],
  pch = pch[group]
)
```

Using this strategy, you can decide for yourself which level of 
deviation between the covariates is still acceptable, while increasing
the difference between sets with regard to the independent variable. 

The function `select_stimuli()` tends to favor the similarity with regard
to the covariates, and tries to maximize differences with regard to the 
independent variable as long as the covariates remain similar.

## Balancing a categorical variable

The function `select_stimuli()` also allows to balance the frequency of 
a categorical variable between stimulus sets via the argument `balance`.
This argument takes as input the names of the columns that should be 
balanced across sets. 

In the following examples, I repeat the four cases from above and use
balance the image category across sets each time.

### Anticlustering

Create sets that are maximally similar with regard to valence, arousal
and beauty from a random preselection (i.e., classical anticlustering),
and the category is balanced between sets:

```{r}
# Preselect 200 random stimuli
preselection <- oasis[sample(nrow(oasis), size = 200), ]
# Create character vector with the relevant covariates:
stimuli <- select_stimuli(
  preselection,
  equalize = c("Valence_mean", "Arousal_mean", "beauty_mean"),
  design = 3,
  balance = "Category"
)
# Check out that the category is balanced:
table(stimuli$SET, stimuli$Category)
# Check M/SD by set
mean_sd_tab(stimuli[, c("Valence_mean", "Arousal_mean", "beauty_mean")], stimuli$SET)
```

Note that by including an additional restriction (i.e., balancing out a 
variable between sets), we make it harder for `anticlust` to find sets 
that are balanced on the covariates. The results are expected to be 
better if no such restriction is included. However, the results in this
case are still pretty good. 

This is a general rule: More restrictions are less easy to satisfy.
If we additionally require that one variable differs between sets, it 
becomes even harder to balance the covariates between sets. However, 
`anticlust` will give its best!

### Min-Max Anticlustering

Create sets that differ on valence, are similar with regard to arousal,
and the category is balanced between sets:

```{r}
# Preselect 200 random stimuli
preselection <- oasis[sample(nrow(oasis), size = 200), ]
stimuli <- select_stimuli(
  preselection,
  split_by = "Valence_mean", 
  equalize = "Arousal_mean",
  design = 3,
  balance = "Category"
)
# Check out that the category is balanced:
table(stimuli$SET, stimuli$Category)
# Check M/SD by set
mean_sd_tab(stimuli[, c("Valence_mean", "Arousal_mean")], stimuli$SET)
```

### Subset Anticlustering

Create sets that are maximally similar with regard to valence, arousal
and beauty. Extract three sets with 50 items each from the entire pool
of 900 items; the category is balanced between sets:

```{r}
stimuli <- select_stimuli(
  oasis,
  equalize = c("Valence_mean", "Arousal_mean", "beauty_mean"),
  design = 3,
  balance = "Category",
  n = 50
)
# Check out that the category is balanced:
table(stimuli$SET, stimuli$Category)
# Check M/SD by set
mean_sd_tab(stimuli[, c("Valence_mean", "Arousal_mean", "beauty_mean")], stimuli$SET)
```

### Subset Min-Max Anticlustering

Create sets that are maximally similar with regard to arousal and 
beauty, maximize difference with regard to valence. Extract three sets 
with 50 items each from the entire pool of 900 items; the category is 
balanced between sets:

```{r}
stimuli <- select_stimuli(
  oasis,
  split_by = "Valence_mean",
  equalize = c("Arousal_mean", "beauty_mean"),
  design = 3,
  balance = "Category",
  n = 50
)
# Check out that the category is balanced:
table(stimuli$SET, stimuli$Category)
# Check M/SD by set
mean_sd_tab(stimuli[, c("Valence_mean", "Arousal_mean", "beauty_mean")], stimuli$SET)
```

## References

Brielmann, A. A., & Pelli, D. (2019). Intense beauty requires intense 
pleasure. *Frontiers in Psychology, 10*, 1--17.

Kurdi, B., Lozano, S., & Banaji, M. R. (2017). Introducing the Open 
Affective Standardized Image Set (OASIS). *Behavior Research Methods, 
49*, 457--470.

Papenberg, M., & Klau, G. W. (2019, October 30). Using anticlustering 
to partition a stimulus pool into equivalent parts. 
https://doi.org/10.31234/osf.io/3razc. Preprint submitted for 
publication.
