---
title: "Using the R package anticlust for stimulus selection in experiments"
output: 
  rmarkdown::html_vignette:
    number_sections: false
    df_print: kable
author: Martin Papenberg
vignette: >
  %\VignetteIndexEntry{Using the R package anticlust for stimulus selection in experiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tableone)
# define a method for objects of the class data.frame
knit_print.matrix = function(x, ...) {
    res = paste(c("", "", kable(x, row.names = TRUE)), collapse = "\n")
    asis_output(res)
}
# register the method
registerS3method("knit_print", "matrix", knit_print.matrix)

```

*Note that I rewrote this vignette in November 2025. The information gathered in this vignette here was quite dated previously, and had a different focus from what I now envision `anticlust` to be used for*

This tutorial shows you how to use the `R` package `anticlust` for stimulus selection in (psychological) experiments. All code can easily be reproduced via Copy & Paste. The tutorial discusses the following functionalities:

- Subset selection from a larger base of stimuli
- Divide the selected stimuli into similar groups of stimuli (via anticlustering)

To follow the code in this tutorial, load the `anticlust` package first:

```{r}
library(anticlust)
```

For the examples in this document, we use norming data for a stimulus set provided by Brunel et al. (2025). It is available when the package `anticlust` is loaded: 

```{r}
# look at the head of the data
head(brunel2025)
```

```{r, echo = FALSE}
cols <- toString(paste0("\`", names(brunel)[3:6], "\`"))
```

The item pool consists of 403 German words, given in the column `target_word`. For their experiments, Brunel et al. investigated how often for a `sentence` the target word is identified (given in the column `percentage_target_word`). 

## Step 1: Subset selection

In many cases, researchers are provided with a relatively large set of stimuli, but only need a subset of the entire pool of stimuli. In many cases, the subset selection is done entirely based on substantive considerations -- there may just be a reason to select some stimuli and not some others. In this case, you can use your subset and procede to divide it into subsets immediately, using anticlustering (see below). 

In some cases, however, researchers may be indifferent or not entirely fixated as to which stimuli to select and may prefer a data driven approach. In the following, I present one such data driven approach, based on maximizing the dispersion in the selected subset. The dispersion is defined as the minimum distance between two stimuli; maximizing the dispersion will ensure that we select stimuli across the entire space of the data describing the stimuli. We can use the `anticlustering()` for this purpose, but note that we actually do not perform anticlustering, which would be a grouping of the stimuli, but we exploit the `K` argument to perform a subset selection. Here, I select stimuli based on the valence and arousal rating of the target words:

```{r}

variables <- brunel2025[, c("valence_target_word", "arousal_target_word")]

N <- nrow(variables)
N_subset <- 100

groups <- anticlustering(
  variables,
  K = c(N_subset, rep(1, N-N_subset)), # crucial that K has this form, for subset selection
  objective = "dispersion",
  method = "local-maximum"
)
groups
```

For this kind of subset selection, it is crucial to define the input for the `K` argument as above, and to select `objective = "dispersion"`. The `groups` vector is the output of the anticlustering function. The selection is coded with 1, so we might want to do some recoding: 

```{r}
selected <- (groups == 1) +1 # 2 = selected, 1 = not selected
table(selected)
```

I can graphically display the selection: 

```{r}

colors <- c("#a9a9a9", "#df536b")
cex <-  c(0.3, 1.2)
pch <- c(19, 15)

plot(
  variables,
  col = colors[selected],
  cex = cex[selected],
  pch = pch[selected]
)

```

The selection is given by the red dots. As we can see, maximizing the dispersion for subset selection ensures that we select stimuli that are representative of the initial sample of stimuli. I think this can be desirable in many applications, but users should decide if this is what they want.

I assume it might also be of interest to select a certain amount of stimuli from different categories. In the current case, we might want to select the same number of words with negative, neutral and positive emotionality. Our first selection did not satisfy this strict condition:

```{r}
table(brunel2025$target_word_emotionality[selected == 2])
```

So we might perform subset selection for each of these word categories separately. Here I assume that I want to select 20 words of each emotionality:

```{r}
N_subset <- 20

negative <- subset(brunel2025, target_word_emotionality == "Negative")
negative$selected <- anticlustering(
  negative[, c("valence_target_word", "arousal_target_word")],
  K = c(N_subset, rep(1, nrow(negative)-N_subset)), 
  objective = "dispersion",
  method = "local-maximum"
)

neutral <- subset(brunel2025, target_word_emotionality == "Neutral")
neutral$selected <- anticlustering(
  neutral[, c("valence_target_word", "arousal_target_word")],
  K = c(N_subset, rep(1, nrow(neutral)-N_subset)), 
  objective = "dispersion",
  method = "local-maximum"
)

positive <- subset(brunel2025, target_word_emotionality == "Positive")
positive$selected <- anticlustering(
  positive[, c("valence_target_word", "arousal_target_word")],
  K = c(N_subset, rep(1, nrow(positive)-N_subset)), 
  objective = "dispersion",
  method = "local-maximum"
)

# Combine in one data frame again:
combined <- rbind(negative, neutral, positive)
combined$selected <- (combined$selected == 1) + 1 # recoding
```

Again, I can illustrate the selection graphically:

```{r}

colors <- c("#a9a9a9", "#df536b")
cex <-  c(0.3, 1.2)
pch <- c(19, 15)


plot(
  combined[, c("valence_target_word", "arousal_target_word")],,
  col = colors[combined$selected],
  cex = cex[combined$selected],
  pch = pch[combined$selected]
)

```

This still looks good, and I ensured an equal number of stimuli from each emotionality category:

```{r}
table(combined$target_word_emotionality[combined$selected == 2])
```

Of course, we can use more than just the two variables (valence and arousal) as a basis for the subset selection. For the purpose of graphical display, I selected two numeric variables, but I could for example also use categorical variables (defined as factors). 

## Step 2: Create stimulus sets via anticlustering

Now I use my selection fo 60 stimuli as input for anticlustering. First, let's actually select the stimuli: 

```{r}
stimuli <- subset(combined, selected == 2)
nrow(stimuli)
table(stimuli$target_word_emotionality)
```

Assume I want 4 equal-sized groups of stimuli for my experiment. Each list should have 5 negative, 5 neutral, and 5 positive words. Moreover, I want to balance the numeric variables valence, arousal, and percent correct between sets. We can for example call anticlustering like this: 

```{r}
balancing_variables <- c("target_word_emotionality", "percentage_target_word", "valence_target_word", "arousal_target_word")
stimuli$Set <- anticlustering(
  stimuli[, balancing_variables],
  K = 4, 
  standardize = TRUE,
  method = "local-maximum", 
  repetitions = 10
)
```

I use the R package `tableone` to create an overview of the desctiprive statistics of all variables: 

```{r}
CreateTableOne(balancing_variables, strata = "Set", data = stimuli)
```

This seems to have worked quite well. Each set has the appropriate number of words from each emotion category. Note that this was not guaranteed: If we want to make sure that a categorical variable has strict priority (before the numeric variables), we can use the `categories` argument instead of including the categorical variable as part of the first argument: 

```{r}
numeric_variables <- c("percentage_target_word", "valence_target_word", "arousal_target_word")
stimuli$Set2 <- anticlustering(
  stimuli[, numeric_variables],
  K = 4, 
  standardize = TRUE,
  method = "local-maximum", 
  repetitions = 10,
  categories = stimuli$target_word_emotionality
)
CreateTableOne(balancing_variables, strata = "Set2", data = stimuli)

```

In the vignette "Using categorical variables with anticlustering", I explore in more detail when it might make sense to use the `categories` argument or not. 

There is some considerations that I might have when doing anticlustering. Note that I set some arguments that I think are useful, which are however not the default: I generally recommend to always use `standardize = FALSE`, especially when combining numeric and categorial variables. Using `method = "local-maximum"` is a potential improvement over the default `method = "exchange"`; using several `repetitions` has the potential to improve results even further. 

We might also want to use a different `objective`, which is the argument that controls how the similarity of stimulus sets is computed. If, for example we use `objective = "kplus"`, we specifically require that means and standard deviations are similar: 

```{r}
numeric_variables <- c("percentage_target_word", "valence_target_word", "arousal_target_word")
stimuli$Set3 <- anticlustering(
  stimuli[, numeric_variables],
  K = 4, 
  standardize = TRUE,
  method = "local-maximum", 
  repetitions = 10,
  categories = stimuli$target_word_emotionality, 
  objective = "kplus"
)
CreateTableOne(balancing_variables, strata = "Set3", data = stimuli)

```

As we can see, standard deviations are now more evenly matched. We might also focus on mean values alone (using `objective = "variance"`), but I generally do not recommend this unless you are very sure what you are doing. 

### Other considerations

Sometimes, we would like to ensure that within each level of a categorical variable -- such as word emotionality -- sets are balanced. Note that this is not achieved using the above calls to anticlustering. First, let's look at the descriptive statistics across groups for each word type separately, from our previous anticlustering assignment:

```{r}
mean_sd_tab(stimuli[, numeric_variables], groups = paste0(stimuli$target_word_emotionality, " - ", "Set ", stimuli$Set3))
```

There are quite strong discrepancies in mean values of negative words between sets. 

```{r}
stimuli$Set4 <- anticlustering(
  stimuli[, numeric_variables],
  K = 4, 
  standardize = TRUE,
  method = "local-maximum",
  blocks = stimuli$target_word_emotionality, 
  objective = "kplus"
)
mean_sd_tab(stimuli[, numeric_variables], groups = paste0(stimuli$target_word_emotionality, " - ", "Set ", stimuli$Set4))

```

These results look better. 

---

**References**


Papenberg, M., & Klau, G. W. (2021). Using anticlustering to partition data sets into equivalent parts. *Psychological Methods, 26*(2), 161–174. https://doi.org/10.1037/met0000301

Schaper, M. L., Kuhlmann, B. G., & Bayen, U. J. (2019a). Metacognitive expectancy effects in source monitoring: Beliefs, in-the-moment experiences, or both? Journal of Memory and Language, 107, 95–110. https://doi.org/10.1016/j.jml.2019.03.009

Schaper, M. L., Kuhlmann, B. G., & Bayen, U. J. (2019b). Metamemory expectancy illusion and schema-consistent guessing in source monitoring. Journal of Experimental Psychology: Learning, Memory, and Cognition, 45, 470. https://doi.org/10.1037/xlm0000602
