---
title: "Using the R package anticlust for stimulus selection in experiments"
output: 
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# define a method for objects of the class data.frame
knit_print.matrix = function(x, ...) {
    res = paste(c("", "", kable(x, row.names = TRUE)), collapse = "\n")
    asis_output(res)
}
# register the method
registerS3method("knit_print", "matrix", knit_print.matrix)


my_scatterplot <- function(
  x, y = NULL,
  xlab = "Valence", 
  ylab = "Arousal", 
  cex.lab = 1.2, 
  cex.axis = 1.2,
  cex = 1,
  frame.plot = FALSE,
  las = 1,
  ...
) {
  plot(
    x, y, 
    xlab = xlab, 
    ylab = ylab, 
    cex.lab = cex.lab, 
    cex.axis = cex.axis, 
    cex = cex, 
    frame.plot = frame.plot,
    las = las,
    ...
  )
}


```

This tutorial teaches you how to use the `R` package `anticlust` for 
stimulus selection. All code can easily be reproduced via Copy & Paste 
if the development version of the `anticlust` package is installed. The 
tutorial discusses the following functionalities:

- Match similar individual stimuli based on covariates of interest
- Minimize differences between stimulus sets with regard to some variables
- Maximize differences between stimulus sets with regard to some variables
- Balance the occurrence of a categorical variable between stimulus sets

First, to follow the code, install the package `anticlust` and load it:

```{r, eval = FALSE}

if (!"remotes" %in% rownames(installed.packages())) {
  install.packages("remotes")
}
remotes::install_github("m-Py/anticlust")

```

```{r}
library(anticlust)
```

For the examples in this document, we use norming data for the open stimulus pool OASIS
(Brielmann & Pelli, 2019; Kurdi, Lozano, & Banaji, 2017). We load the 
data provided by Brielmann and Pelli (2019) from the internet and save 
the data as an `R` data frame: 

```{r}
oasis <- read.csv("https://raw.githubusercontent.com/aenneb/OASIS-beauty/master/means_per_image.csv")
```

Let's check out the data: 

```{r}
head(oasis)
nrow(oasis)
```

The OASIS data set contains norming data for 900 photos on 
arousal, valence, and beauty; so we are interested in the columns 
`Valence_mean`, `Arousal_mean`, `beauty_mean` to select stimulus sets.

In a first example, I create groups of stimuli that are matched on arousal
but are dissimilar on valence. To this end, I recode the numeric variable
`Valence_mean` into three categories, representing negative, neutral and 
positive valence. The following scatter plot illustrates the boundaries that I 
set to define negative, neutral and positive valence.


```{r}

oasis <- within(oasis, {
  valence <- ifelse(Valence_mean <= 2.5, "Valence-low", NA)
  valence <- ifelse(Valence_mean > 3.2 & Valence_mean < 4.2, "Valence-neutral", valence)
  valence <- ifelse(Valence_mean > 5.2, "Valence-high", valence)
})

selected <- subset(oasis, !is.na(valence))

# Illustrate final stimulus selection
was_selected <- oasis$item %in% selected$item
group <- rep(1, nrow(oasis))
group[was_selected] <- to_numeric(selected$valence) + 1


colors <- c("#a9a9a9", "#df536b", "#61d04f", "#2297e6")
cex <-  c(0.1, rep(1.1, 3))
pch <- c(4, 15, 17, 19)
my_scatterplot(
  oasis[, c("Valence_mean", "Arousal_mean")],
  col = colors[group],
  cex = cex[group],
  pch = pch[group]
)

```

Next, we conduct a matching: We find triplets from the three different groups 
that are as similar as possible with regard to arousal:

```{r}

# Match the conditions based on covariates
selected$matches <- matching(
  selected$Arousal_mean, 
  match_between = selected$valence,
  match_extreme_first = FALSE
)

selected <- subset(selected, !is.na(matches))

# check out some of the best match:
subset(selected, matches == 1)
subset(selected, matches == 2)

# Select the 25 best matches:
final_selection <- subset(selected, matches < 30)


# Check quality of the selection:
mean_sd_tab(
  subset(final_selection, select = c(Valence_mean, Arousal_mean)), 
  final_selection$valence
)


# Illustrate final stimulus selection
was_selected <- oasis$item %in% final_selection$item
group <- rep(1, nrow(oasis))
group[was_selected] <- to_numeric(final_selection$valence) + 1


colors <- c("#a9a9a9", "#df536b", "#61d04f", "#2297e6")
cex <-  c(0.1, rep(1.1, 3))
pch <- c(4, 15, 17, 19)
my_scatterplot(
  oasis[, c("Valence_mean", "Arousal_mean")],
  col = colors[group],
  cex = cex[group],
  pch = pch[group]
)

```


## References

Brielmann, A. A., & Pelli, D. (2019). Intense beauty requires intense 
pleasure. *Frontiers in Psychology, 10*, 1--17.

Kurdi, B., Lozano, S., & Banaji, M. R. (2017). Introducing the Open 
Affective Standardized Image Set (OASIS). *Behavior Research Methods, 
49*, 457--470.

Papenberg, M., & Klau, G. W. (2019, October 30). Using anticlustering 
to partition a stimulus pool into equivalent parts. 
https://doi.org/10.31234/osf.io/3razc. Preprint submitted for 
publication.
